GSS Education Analysis (Working Title)

### 2023-03-16 outputs/paper/paper.qmd
update paper

@@

The United States is built upon individualism, which trickles into their definitions of financial security, happiness, and overall fulfillment and satisfaction in life. Well-being of individuals within the United States therefore, is tracked and monitored through a variety of different factors. The aim of this investigation was to determine whether higher educational achievement influenced an individuals perceptions of social class, degree of financial satisfaction and job satisfaction. These factors were chosen because of their significance with regards to the American Dream, a concept that is built off a linear pattern of behavior in terms of economic security. Hustle culture is popular within the United States as an extension of the American Dream, because the idea that an individual must work hard in order to reap the benefits in terms is what drives culture within America.

According to Economic Security and the American Dream, the rewards of the American Dream included owning a home, having access to quality health care, having a job that enables an individual to support their family, and having a secure and stately retirement [@citerankhirschl]**.** These are the fundamentals in terms of how Americans generally define their well-being in society. However, no matter how linear an American's personal development or educational journey may be, it is simply impossible to be able to guarantee the rewards that the American Dream has to offer. But the ideals of living in the United States, has a powerful impact on culture, and how individuals view themselves and their position within society, but how they fare in comparison to those around them.

The assumption in general is a higher degree of education translates so an overall greater quality of life or well-being, greater job satisfaction, higher social class, etc. However, we found that this was not necessarily the case. Overall out of our participants that completed the survey, the largest group had only completed high school, and the second largest group had completed a bachelor's degree. The level of education was compared to a few variables: overall happiness, self-ascribed social class, job satisfaction, and financial ranking against other families in the US.

@@

Researchers have investigated the correlation between education and happiness, with several studies indicating that a higher level of education is positively associated with greater levels of happiness. This relationship may be explained by the fact that education can impact one's self-ascribed social class, which in turn can influence their notion of happiness. Higher education can provide individuals with better job opportunities, higher income, better access to resources, and greater social mobility, leading to a higher sense of social status. As a result, these individuals may experience greater life satisfaction, which contributes to their overall sense of happiness and security.

In a 2020 article published in the Journal of Personality and Social Psychology, authors Siyu Yu and Steven L. Blader examined the impact of status and social class on subjective well-being (SWB). They found that social class addresses fundamental psychological needs, such as autonomy, mastery, respect, and occupational prestige, and proposed that individuals in higher social classes experience greater levels of status and power, leading to higher SWB [@citeyublader]Conversely, individuals in lower social classes feel lower levels of status and power, resulting in lower levels of SWB. As shown in @fig-4 we found that individuals holding Bachelors and Graduate Degrees are proportionally more satisfied with their financial situation, with the large majority selecting "Pretty Well Satisfied" and "More or Less Satisfied" than those selecting "Not Satisfied at All". Both the research and data point to the link between higher social class and greater levels of self-ascribed satisfaction and happiness.

Overall, there seems to be a significant relationship between education, social class, and one's perceived well-being. Education can serve as a key factor in determining an individual's social class as it unlocks opportunities for resources and mobility, which in turn can impact their sense of status and overall sense of happiness. At the same time, while higher levels of education may be generally associated with greater levels of happiness, this relationship may not always be a cause and effect correlation. Subjective well-being may be further influenced by other factors such as personal values, cultural background, and social support. It is important to consider these important nuances when examining the data in order to develop a more comprehensive understanding of how variables interact and impact individuals' lives. This will be further explored in the weaknesses, ethics, and bias section of this paper.

@@

Financial satisfaction as a result of hard work is the core principle within the American Dream, and as a result, Americans are known for their work culture. Researchers have tried to investigate the differences between work culture within Europe as well as within the United States and have narrowed it down to a variety of factors that contribute to the culture as it is today. Despite the fact that the economies and societal frameworks of Europe and America are very similar, the approach to work within the United States is very different. The number of hours worked per year for the average American was about 17% higher than the hours per year for the average European [@citeeuropevacay]. For this reason, financial satisfaction and social class are very prevalent within the American definition of well-being. Job satisfaction ties very closely this idea in the pursuit of the American Dream.

The second largest group of participants who had completed the survey had completed a bachelor's degree. In general, while a higher level of education improves access to better job opportunities, the process is not as linear as it may seem. According to @fig-1, while the number of individuals who have a graduate degree is not as high as the group that has completed a bachelor's degree, the reported financial ranking against other families was proportionately very similar for both groups. The majority for both groups claimed to be average or above average. There was still a significant number of individuals who had chosen below average, even for the group with the highest level of education.

A similar notion exists when comparing an individuals level of education with their job satisfaction. According to @fig-2, while most participants with an education level above high school, the majority group of individuals with a bachelor's or graduate degree still consists of those that chose "Moderately Satisfied", "A Little Dissatisfied", and "Very Dissatisfied". This points towards the unattainability of the American Dream, in that despite the greater level education and increased access that comes with it, it does not guarantee increased job satisfaction, or a higher financial satisfaction. According to @fig-3, the two highest levels of education still make up a significant portion of the "More or Less Satisfied" and the "Not Satisfied at All'' groups.

@@

There are several limitations to this paper. One weakness is that our insights rely on data from the General Social Survey (GSS), which is a cross-sectional survey that collects data at a single point in time. While it can be useful for exploring relationships between variables and generating hypotheses, it cannot establish causality between variables, has limited ability to assess change over time, and has potential for selection bias if participants are not representative of the population of interest [@citestrengthsweaknesses]**.** Due to the nature of the data being self-reported responses, it may not fully capture the dynamic nature of the ways in which various levels of education affect one's overall well-being. GSS responses are subject to influence by external factors such as recall bias, in which participants may have left out details when reporting about their subjective experiences.

Another potential ethical concern is that our paper may perpetuate the notion that higher education is the only solution to achieving happiness and success, disregarding the presence of systemic inequalities such as racism and economic inequality that can limit access to education and social mobility. Moreover, the sample survey in the GSS may not fully represent the experiences of more marginalized groups, which point to the need for further research and survey strategies that incorporate more diverse perspectives.

In sum, while the paper provides valuable insights on the relationship between education and self-perceived happiness, it is important to consider the limitations, ethical concerns, and potential biases in the data and interpretation of our findings.

@@

More research into this topic is important as it helps us gain a more substantial understanding of American ideals and how certain indicators may influence the rate at which individuals are achieving them, or not achieving them. Further research into this topic should explore more specific questions about attitudes towards work, type of work, income, and comparing respondent situations to close acquaintances to better understand American perspectives. We have provided a supplementary survey in @sec-appendix that would further build on the US GSS data in relation to our topic. Additionally, future research should be conducted over a period of time using the same survey and methodology to better obtain information that is verifiable and encompasses a larger population of the United States.

>>>>

By looking at @fig-4 we can see that those with graduate and bachelor's degrees are much more likely to describe themselves as "upper class" than "lower class." Interestingly, in all levels of education, most respondents ranked themselves as either "working class" or "middle class" this could be influenced by American idealism of "hard-work" or how respondent's see themselves among their peers which is also subjective. Additionally, only a very small number of those with associate or junior college degrees and lower education levels describe themselves as "upper class," ranging from 0 to 10% of respondents from these groups.

@@

The United States is built upon individualism, which trickles into their definitions of financial security, happiness, and overall fulfillment and satisfaction in life. Well-being of individuals within the United States therefore, is tracked and monitored through a variety of different factors. The aim of this investigation is to determine whether higher educational achievement influences respondent\'s perceptions of their social class, degree of financial satisfaction and job satisfaction. These factors were chosen because of their significance with regards to the American Dream, a concept that is built off a linear pattern of behavior in terms of economic security. Hustle culture is popular within the United States as an extension of the American Dream, because the idea that an individual must work hard in order to reap the benefits in terms is what drives culture within America.

According to Economic Security and the American Dream, the rewards of the American Dream included owning a home, having access to quality health care, having a job that enables an individual to support their family, and having a secure and stately retirement [@citerankhirschl]**.**These are the fundamentals in terms of how Americans generally define their well-being in society. However, no matter how linear an American\'s personal development or educational journey may be, it is simply impossible to be able to guarantee the rewards that the American Dream has to offer. But the ideals of living in the United States, has a powerful impact on culture, and how individuals view themselves and their position within society, but how they fare in comparison to those around them.

The assumption in general is a higher degree of education translates to an overall greater quality of life or well-being, greater job satisfaction, higher social class, etc. However, we found that this was not necessarily the case. Overall, out of our participants that completed the survey, the largest group had only completed high school, and the second largest group had completed a bachelor\'s degree. The level of education was compared to a few variables: self-ascribed social class, financial satisfaction,  job satisfaction, and financial ranking against other families in the US.

@@

However, in some cases, the positive correlation between higher education and overall subjective well-being is more prominent - resulting in significant tension in our data of consideration. Researchers have investigated the relationship between education and one\'s subjective well-being, with several studies indicating that a higher level of education may be positively associated with greater levels of happiness. This relationship may be explained by the fact that education can impact one\'s self-ascribed social class, which in turn can influence their notion of happiness. Higher education is assumed to provide individuals with better job opportunities, higher income, better access to resources, and greater social mobility - placing people on the path towards a successful American Dream and resulting in a higher sense of one\'s social status. As a result, these individuals might experience greater life satisfaction, which contributes to their overall sense of happiness and security.

In a 2020 article published in the Journal of Personality and Social Psychology, authors Siyu Yu and Steven L. Blader examined the impact of status and social class on subjective well-being (SWB). They found that often times social class addresses fundamental psychological needs, such as autonomy, mastery, respect, and occupational prestige, and they proposed that individuals in higher social classes experience greater levels of status and power, leading to higher SWB **[@citeyublader].** Conversely, individuals in lower social classes feel lower levels of status and power, resulting in lower levels of SWB. This insight is evident in @fig-3, where we found that individuals holding Bachelors and Graduate Degrees are proportionally more satisfied with their financial situation, with the large majority selecting \"Pretty Well Satisfied\" and \"More or Less Satisfied\" than those selecting \"Not Satisfied at All\". Out of the 426 responses indicating a \"Pretty Well\" financial satisfaction, 69% of responses were individuals holding Bachelor and Graduate degrees while only 31% held Associate/Junior college, highschool, and less than highschool levels of education. In comparison, those who only completed up until the high school level of education have the most responses of \"Not Satisfied at All\". 70% out of the 390 \"Not Satisfied At All\" responses were individuals who completed highschool education or lower, while only 30% of those holding Bachelor and Graduate degrees indicated a low financial satisfaction.

In addition, when looking at @fig-4, we found that there are more individuals holding Bachelor and Graduate degrees who have identified themselves as Upper and Middle class in proportion to those with lower levels of education. 89% of those who selected \"Upper Class\" and 68% of \"Middle Class\" held Bachelor and Graduate degrees. Among the 676 responses for \"Working Class\", 471 of those held high school level of education and lower, making up 70% of the responses. In sum, we found that both the research and data point to a link between higher social class and greater perceived levels of life satisfaction.

Overall, there seems to be a significant relationship between the data in education, social class, and one\'s perceived well-being. Education may serve as an important factor in determining an individual\'s social class as it can unlock opportunities for resources and mobility, which in turn can positively impact their sense of status and overall happiness. At the same time, while higher levels of education may be generally associated with greater levels of overall well-being, this relationship may not always be a cause and effect correlation. Subjective well-being may be further influenced by other factors such as personal values, cultural background, and social support. It is important to consider these important nuances when examining the data and making assumptions in order to develop a more comprehensive understanding of how variables interact and impact individuals\' lives. This will be further explored in the weaknesses, ethics, and bias section of this paper.

@@

Financial satisfaction as a result of hard work is the core principle within the American Dream, and as a result, Americans are known for their work culture. Researchers have tried to investigate the differences between work culture within Europe as well as within the United States and have narrowed it down to a variety of factors that contribute to the culture as it is today. Despite the fact that the economies and societal frameworks of Europe and America are very similar, the approach to work within the United States is very different. The number of hours worked per year for the average American was about 17% higher than the hours per year for the average European. [@citeeuropevacay] For this reason, financial satisfaction and social class are very prevalent within the American definition of well-being. Job satisfaction ties very closely this idea in the pursuit of the American Dream.

The second largest group of participants who had completed the survey had completed a bachelor's degree. In general, while a higher level of education improves access to better job opportunities, the process is not as linear as it may seem. According to @fig-1, while the number of individuals who have a graduate degree is not as high as the group that has completed a bachelor\'s degree, the reported financial ranking against other families was proportionately very similar for both groups. The majority for both groups claimed to be average or above average. There was still a significant number of individuals who had chosen below average, even for the group with the highest level of education.

A similar notion exists when comparing an individuals level of education with their job satisfaction. According to @fig-2,as the education level rises so does the number of people who are proportionately more satisfied with their jobs. However, job satisfaction is fleeting, it is simply a reflection of the current state, and not an overall indication of cumulative job satisfaction. According to Figure 3, financial satisfaction rises with an increased level of education, but the group that claims dissatisfaction is a significant number in the bachelor\'s and graduate degree groups. It remains a minority, but it is significant nonetheless. This points towards the un-attainability of the American Dream, in that despite the greater level education and increased access that comes with it, it does not guarantee increased job satisfaction, or a higher financial satisfaction.

@@

There are several limitations to this paper. One weakness is that our insights rely on data from the General Social Survey (GSS), which is a cross-sectional survey that collects data at a single point in time. Although it can be helpful in investigating associations between variables and developing hypotheses, the GSS is not capable of establishing causation between variables and has potential for selection bias if participants are not representative of the population of interest [@citestrengthsweaknesses]. In addition, due to changes in the questions asked over time, the GSS has limited capacity to make exact comparisons - as in the data analyzed in our study, which was collected during a period when the survey was conducted entirely online, which necessitated modifications to the question format in order to accommodate the new platform. Above all, due to the nature of the data being self-reported responses, it may not fully capture the dynamic and complicated nature of the ways in which various levels of education affect one\'s overall well-being. The General Social Survey responses are subject to influence by a myriad of external factors including recall bias, in which participants may have left out important details when reporting about their subjective experiences.

Another potential ethical concern is that our paper may inadvertently reinforce the concept of the American Dream, which suggests that higher education is a key factor to achieving happiness and success in life. This concern is particularly relevant in @sec-social-class, where we analyze the data as-is without taking into account the presence of systemic inequalities such as racism and economic inequality that can significantly limit access to one\'s education and social mobility. Moreover, the sample survey in the GSS may not fully represent the experiences of more marginalized groups, which point to the need for further research and survey strategies that incorporate more diverse perspectives.

In sum, while our paper provides valuable insights on the relationship between education and one\'s perceived well-being, it is important to consider the limitations, ethical concerns, and potential biases in the data and interpretation of our insights and findings.

@@

More research into this topic is important as it helps us gain a more substantial understanding of American ideals and how certain indicators may influence the rate at which individuals are achieving them, or not achieving them. Further research into this topic should explore more specific questions about attitudes towards work, type of work, income, and comparing respondent situations to close acquaintances to better understand American perspectives. We have provided a supplementary survey in @sec-appendix that would further build on the US GSS data in relation to our topic. Additionally, future research should be conducted over a period of time using the same survey and methodology to collect obtain information that is verifiable and encompasses a larger population of the United States.
